---
title: "Class 5 Data Vis with GGPLOT"
author: "Rachel Field (PID A69042948)"
format: pdf
---

Today we are playing with plotting and graphics in R. 

There are a lot of ways to make cool figures in R.

There is "base" R graphics (`plot()`, `hist()`. `boxplot()`, etc.)

There is also add-on packages like **ggplot**

```{r}
head(cars, 3)
```
Let's plot this with "base" R.

```{r}
plot(cars)
```
```{r}
head(mtcars)
```
Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=10)
```
```{r}
hist(mtcars$mpg)
```


##GGPLOT

The main function of the ggplot2 package is **ggplot()**. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`

> **N.B.** I never want tp run `install.packages()` in my quarto source document!!!

```{r}
library(ggplot2)
ggplot(cars) + aes(x=speed, y=dist) + geom_point()
```

```{r}
ggplot(cars) + aes(x=speed) + geom_histogram()
```
Every ggplot needs at least 3 things:
- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

>For simple canned graphs "base" R is nearly always faster

###Adding more layers

Let's add a line and a title, subtitle, and caption as well as custom axis labels.

```{r}
ggplot(cars) +
  aes(x=speed, y = dist) +
  geom_point() +
  geom_line()
```


```{r}
ggplot(cars) +
  aes(x=speed, y = dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs (title = "Speed versus Distance", subtitle = "Cars", x= "Speed (MPH)", y = "Distance (Feet)", caption = "caption!") +
  theme_bw()
```
##Plot some expression data

Read data file from onlone url

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
>Q1: How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset.

```{r}
nrow(genes)
colnames(genes)
ncol(genes)
```


>Q2: How many upregulated genes are there?

```{r}
sum(genes$State=="up")
table(genes$State)
round( table(genes$State)/nrow(genes) * 100, 2 )
```

```{r}
ggplot(genes) + 
  aes(x=Condition1, y = Condition2, color=State) +
  geom_point()
```
```{r}
p<-ggplot(genes) + 
  aes(x=Condition1, y = Condition2, color=State) +
  geom_point() +
  scale_color_manual(values = c("purple", "orange", "blue"))

p+labs(title="look at me")
```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State, label=Gene)+
  geom_point()+
  scale_color_manual(values =c("blue", "gray", "red"))+
  geom_text()
```

```{r}
library(ggrepel)
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State, label=Gene)+
  geom_point()+
  scale_color_manual(values =c("blue", "gray", "red"))+
  geom_text_repel(max.overlaps = 60) +
  theme_bw()
```

##Going Further

Playing with some different layers and the gapminder dataset.

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```


```{r}
tail(gapminder, 6)
```

A first plot

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y= lifeExp, color=continent) +
  geom_point() +
  facet_wrap(~continent)
```

##Optional plot, whatever we want
```{r}
ggplot(gapminder)+
  aes(x=year, y=lifeExp, color=continent) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~continent)
```

##Other plot from lab sheet

```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

